signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-19 paper.tex
remade tables
more discussion improvments

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \toprule
Criterion name & Property name \\ 
  \midrule
Size & Size~\citep{dawson1972size} \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight~\citep{fasman1976handbook}\\ 
  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
  Polarity & Polarity~\citep{1974granthamamino} \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
   \bottomrule
\end{tabular}
\caption{Properties used in clusterization.} 
\label{tab:aaprop}
\end{table}

@@

\section*{Results}

@@

\begin{table}[ht]
\begin{minipage}{.5\linewidth} 
\centering
\begin{tabular}{l}
  \toprule
Groups \\ 
  \midrule
D, E, H, K, N, Q, R \\ 
   \rowcolor[gray]{0.85}G, P, S, T, Y \\ 
  F, I, L, M, V, W \\ 
   \rowcolor[gray]{0.85}A, C \\ 
   \bottomrule
\end{tabular}
\caption{The best sensitivity (final) encoding.} 
\label{tab:best}
\end{minipage}
\begin{minipage}{.5\linewidth} 
\centering
\begin{tabular}{l}
  \toprule
Groups \\ 
  \midrule
A, E, K, Q, R \\ 
   \rowcolor[gray]{0.85}D, G, N, P, S, T \\ 
  C, H, I, L, M, V \\ 
   \rowcolor[gray]{0.85}F, W, Y \\ 
   \bottomrule
\end{tabular}
\caption{The worst sensitivity encoding.} 
\label{tab:worst}
\end{minipage}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \toprule
Measure & Mean & SD \\ 
  \midrule
AUC & 0.9683 & 0.0024 \\ 
   \rowcolor[gray]{0.85}MCC & 0.8677 & 0.0049 \\ 
  Sensitivity & 0.9406 & 0.0008 \\ 
   \rowcolor[gray]{0.85}Specificity & 0.9269 & 0.0050 \\ 
   \bottomrule
\end{tabular}
\caption{Performance measures for the best encoding. 60 repetitions of cross-validation.} 
\label{tab:perfmeas}
\end{table}

@@

We examined more deeply encodings with the best and the worst sensitivity. In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals. In the best encoding, all charged amino acids, both acids and even weak bases as histidine, also belong to this group. This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding. In the worst encoding, where polar and charged character of the Group 1 is not that explicit, the difference in frequency between regions is also less visible.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \toprule
Software name & AUC & Sensitivity & Specificity & MCC \\ 
  \midrule
signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9424 & 0.9720 & 0.9128 & 0.8859 \\ 
   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9675 & 0.9579 & 0.9771 & 0.9353 \\ 
  PrediSi \citep{2004hillerpredisi} & 0.8959 & 0.9065 & 0.8853 & 0.7919 \\ 
   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.9515 & 0.9673 & 0.9358 & 0.9033 \\ 
  Philius \citep{2008reynoldstransmembrane} & 0.9376 & 0.9533 & 0.9220 & 0.8755 \\ 
   \rowcolor[gray]{0.85}signalHsmm-2010 & 0.9429 & 0.9393 & 0.8761 & 0.8166 \\ 
  signalHsmm-1989 & 0.9455 & 0.9486 & 0.8945 & 0.8440 \\ 
   \bottomrule
\end{tabular}
\caption{Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.} 
\label{tab:bench2010}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \toprule
Software name & AUC & Sensitivity & Specificity & MCC \\ 
  \midrule
signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & 0.9469 & 0.4028 \\ 
  Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
   \rowcolor[gray]{0.85}Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
  signalHsmm-2010 & 0.9340 & 0.7941 & 0.8939 & 0.6526 \\ 
   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & 0.7531 \\ 
   \bottomrule
\end{tabular}
\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
\label{tab:bench2010plas}
\end{table}

\section*{Discussion}

Despite the presence of signal peptide predicting software, there are no exhaustive studies devoted to the analysis of the signal peptide structure and taxonomical variability. Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution.

@@

Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also found new relationships. We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins. The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.

The proposed algorithm recognizing various signal peptides will reduce costs and speed up searching appropriate artificial signal peptides designed for protein secretion in expensive and time-consuming laboratory experiments. Moreover, the algorithm can be applied in recognition any sequence types on the nucleotide and amino acid levels, represented by small sets. Accessible as a stand-alone and browser-based software, its useful both for researches interested in the analysis of signal peptides.

>>>>

\begin{table}[ht]
\centering
\caption{Properties used in clusterization.} 
\begin{tabular}{ll}
  \toprule
Criterion name & Property name \\ 
  \midrule
Size & Size~\citep{dawson1972size} \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight~\citep{fasman1976handbook}\\ 
  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
  Polarity & Polarity~\citep{1974granthamamino} \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
   \bottomrule
\end{tabular}
\label{tab:aaprop}
\end{table}

@@

\section*{Results and discussion}

@@

\begin{table}[ht]
\begin{minipage}{.5\linewidth} 
\centering
\caption{The best sensitivity (final) encoding.} 
\begin{tabular}{l}
  \toprule
Groups \\ 
  \midrule
D, E, H, K, N, Q, R \\ 
   \rowcolor[gray]{0.85}G, P, S, T, Y \\ 
  F, I, L, M, V, W \\ 
   \rowcolor[gray]{0.85}A, C \\ 
   \bottomrule
\end{tabular}
\label{tab:best}
\end{minipage}
\begin{minipage}{.5\linewidth} 
\centering
\caption{The worst sensitivity encoding.} 
\begin{tabular}{l}
  \toprule
Groups \\ 
  \midrule
A, E, K, Q, R \\ 
   \rowcolor[gray]{0.85}D, G, N, P, S, T \\ 
  C, H, I, L, M, V \\ 
   \rowcolor[gray]{0.85}F, W, Y \\ 
   \bottomrule
\end{tabular}
\label{tab:worst}
\end{minipage}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Performance measures for the best encoding. 60 repetitions of cross-validation.} 
\begin{tabular}{lrr}
  \toprule
Measure & Mean & SD \\ 
  \midrule
AUC & 0.9683 & 0.0024 \\ 
   \rowcolor[gray]{0.85}MCC & 0.8677 & 0.0049 \\ 
  Sensitivity & 0.9406 & 0.0008 \\ 
   \rowcolor[gray]{0.85}Specificity & 0.9269 & 0.0050 \\ 
   \bottomrule
\end{tabular}
\label{tab:perfmeas}
\end{table}

@@

We examined more deeply encodings with the best and the worst sensitivity. In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals. In the best encoding, all charged amino acids, both acids and bases (even as weak bases as histidine), also belong to this group. This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding, because is nearly absent in the h-region. In the worst encoding, where polar and charged character of the Group 1 is not that explicit, the difference in frequency between regions is also less visible.

@@

\begin{table}[ht]
\centering
\caption{Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.} 
\begin{tabular}{lrrrr}
  \toprule
Software name & AUC & Sensitivity & Specificity & MCC \\ 
  \midrule
signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9424 & 0.9720 & 0.9128 & 0.8859 \\ 
   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9675 & 0.9579 & 0.9771 & 0.9353 \\ 
  PrediSi \citep{2004hillerpredisi} & 0.8959 & 0.9065 & 0.8853 & 0.7919 \\ 
   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.9515 & 0.9673 & 0.9358 & 0.9033 \\ 
  Philius \citep{2008reynoldstransmembrane} & 0.9376 & 0.9533 & 0.9220 & 0.8755 \\ 
   \rowcolor[gray]{0.85}signalHsmm-2010 & 0.9429 & 0.9393 & 0.8761 & 0.8166 \\ 
  signalHsmm-1989 & 0.9455 & 0.9486 & 0.8945 & 0.8440 \\ 
   \bottomrule
\end{tabular}
\label{tab:bench2010}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
\begin{tabular}{lrrrr}
  \toprule
Software name & AUC & Sensitivity & Specificity & MCC \\ 
  \midrule
signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & 0.9469 & 0.4028 \\ 
  Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
   \rowcolor[gray]{0.85}Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
  signalHsmm-2010 & 0.9340 & 0.7941 & 0.8939 & 0.6526 \\ 
   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & 0.7531 \\ 
   \bottomrule
\end{tabular}
\label{tab:bench2010plas}
\end{table}

\subsection*{Conclusions}

Despite the presence of signal peptide predicting software, there are no exhaustive studies devoted to the analysis of the signal peptide structure and taxonomical variability. Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution. Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also discovered new relationships. We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins. The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.

@@

The flexibility and efficient information recovery makes our model unique among similar software. The signalHsmm not only shows very good performance but also is adjustable to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.

\subsection*{Availability and implementation}
The signalHsmm web-server is avaible under address: \url{http://smorfland.uni.wroc.pl/signalhsmm}.

signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signal.hsmm}.
